Red wine quality EDA by Alfred Wong
========================================================



```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(plyr)
library(ggplot2)
library(gridExtra)
library(tidyr)
library(GGally)
```

```{r load data, echo=FALSE}
setwd('C:/Users/Din/Documents/Projects/Data_science/Udacity_EDA/EDA_project_red_wine/')
rwq <- read.csv('wineQualityReds.csv')
rwq <- as_tibble(rwq)

```

# Introduction

In this study we explore a dataset about wine quality created by Cortez et al.
(2009) in:

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

Available at: 

  [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
  
  [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
  
  [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib

It has with 1599 observations of 12 variables.

Each observation is a bottle of the Portuguese "Vinho Verde" red wine. 
Our main feature here is the quality of the bottle, which is the median of 
integer (0-10) ratings given by at least three experts.

The rest of the features are measurements given by physiochemical tests. (The 
variable *X* shown below is just the index)

Our aim is to explore the data, paving the way for further studies to find out
what features can help predict that a wine will be good.

```{r echo=FALSE, Univariate_Plots}
str(rwq)
summary(rwq)

rwq <- rwq %>%
  select(quality, alcohol, fixed.acidity, volatile.acidity, citric.acid,
         pH, residual.sugar, chlorides, free.sulfur.dioxide, 
         total.sulfur.dioxide, sulphates, density)
```

Here is the overview of the data.

```{r echo=FALSE}
colSums(is.na(rwq))
```

We don't have any NA values.



# Univariate Plots Section

Now let's take a look at the variables one by one.


### *quality*

```{r quality, echo=FALSE}
ggplot(rwq, aes(x = quality)) +
  geom_bar(color = "red")
```

*quality* is the rating of the wine. 3 or more experts rated the wine from 0, 
the worst, to 10, the best, and the median is taken.
All of the values we have are between 3 and 8, and mostly concentrated in 5 and 
6.
This will be the focus of our study, as our aim is to explore the variables 
affecting *quality*.


### *alcohol*

```{r alcohol, echo=FALSE}
ggplot(rwq, aes(x = alcohol)) +
  geom_histogram(binwidth = 0.1, color = "red") 
```

*alcohol* is the alcohol content by volume. We see that most of our wines have 
values between 9 and 14, with a mean of 10.42 and a median of 10.20. 


### *fixed.acidity*

```{r fixed.acidity, echo=FALSE}
ggplot(rwq, aes(x = fixed.acidity)) +
  geom_histogram(bins = 50, color = "red")
```

*fixed.acidity* is the amount of tartaric acid present in the wine.
Most wines in our data set have around 8 gram per liter. The distribution have 
a bit of skewedness towards the right.

```{r echo=FALSE}
ggplot(rwq, aes(x = fixed.acidity)) +
  geom_histogram(bins = 50, color = "red")  +
  scale_x_log10()
```

With the log transformation the distribution looks somewhat closer to normal.


### *volatile.acidity*

```{r volatile.acidity, echo=FALSE}
ggplot(rwq, aes(x = volatile.acidity)) +
  geom_histogram(bins = 50, color = "red")
```

*volatile.acidity* is the amount of acetic acid, in grams / liter.
Usually it is associated with bad quality with its vinegar-y smell.
The distribution looks bimodal, with peaks at around 0.4 and 0.6 g/L


### *citric.acid*

```{r citric.acid, echo=FALSE}
ggplot(rwq, aes(x = citric.acid)) +
  geom_histogram(binwidth = 0.01, color = "red")
```

*citric.acid* is the amount of citric acid in the wine, again in grams / liter.
Many of the wines in our dataset has none at all. There are a few spots of 
concentrated frequencies at 0.49, 0.24 and 0.02. This is not a binning problem 
in our plot, as I have matched the binwidth with the data.

As for the other values, we have a few outliers at 1.00, and the rest are 
somewhat uniformly distributed up to around 0.5, and then gradually decrease.


### *pH*

```{r pH, echo=FALSE}
ggplot(rwq, aes(x = pH)) +
  geom_histogram(binwidth = 0.02, color = "red") +
  scale_x_continuous(breaks = seq(2.75, 4, 0.25))
```

The pH value of the wine. The lower the value the more acidic is the wine. 

The distribution is symmetric with a mean of 3.311.

Note that pH value itself is a logarithmic measurement of hydrogen ions, so at 
least chemically speaking, a wine with a pH value of 3 is ten times more acidic
than one with a 4.

However, pH value is not exactly the log of (*fixed.acidity* + 
*volatile.acidity*), as the pH value is affected by both the amount and strength
of acids present, while the acidity values count only the amount of the acids 
concerned.


### *residual.sugar*

```{r residual.sugar, echo=FALSE}
ggplot(rwq, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.1, color = "red")
```

*residual.sugar* is the amount of sugar, in gram / liter, left in the wine after 
the fermentation process ends, or stopped by the winemaker. 

Ceteris paribus, the more residual sugar there is, the sweeter the wine should 
taste. But then the sensation of sweetness is also affected by many other 
factors, such as acidity.

The majority of observations form a bell shape, with the rest forming a long 
tail to the right. The median is 2.2 while the mean is 2.54.

```{r echo=FALSE}
ggplot(rwq, aes(x = residual.sugar)) +
  geom_histogram(bins = 45, color = "red") +
  scale_x_log10()
```

Again with a log scale the distribution looks more symmetric.


### *chlorides*

```{r chlorides, echo=FALSE}
ggplot(rwq, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.005, color = "red")
```

*cholorides* is the amount of sodium chlorides (ie, salt) in the wine, in gram /
liter. In large enough amounts they give the wine a salty taste.
Most of the wines in our sample have less than 0.12 g/L, but some outliers 
have up to 0.6 g/L.

```{r echo=FALSE}
ggplot(rwq, aes(x = chlorides)) +
  geom_histogram(bins = 100, color = "red") +
  scale_x_log10()
```

Here the log transformation makes the main part of the distribution less 
bell-like, but evens out the tails on both sides a little.


### *free.sulfur.dioxide*

```{r free.sulfur.dioxide, echo=FALSE}
ggplot(rwq, aes(x = free.sulfur.dioxide)) +
  geom_histogram(binwidth = 1, color = "red")
```

Sulfur dioxide is a by-product of the fermentation process, but is also added
to the wine by the winemaker. It serves as 1. a anti-microbial agent that 
suppresses harmful yeast and bacteria growth, and 2. an anti-oxidatant to keep
the wine fresh.

*free.sulfur.dioxide* is the amount that exists as molecular SO2, sulfite or 
bisulfite ions, that helps suppress microbes and oxidation. The unit is
mg / liter.

The distribution of *free.sulphur.dioxide* has a tail to the right. But neither
the logarithic or the square root transformation would make it very 
bell-shaped.

```{r echo=FALSE}
ggplot(rwq, aes(x = free.sulfur.dioxide)) +
  geom_histogram(bins = 30, color = "red") +
  scale_x_log10()
```

Like in this log-transformed plot.


### *total.sulfur.dioxide*

```{r total.sulfur.dioxide, echo=FALSE}
ggplot(rwq, aes(x = total.sulfur.dioxide)) +
  geom_histogram(bins = 50, color = "red")
```

Bound sulfur dioxide is the portion that is already attached to other compounds
in the wine, and thus does not work as antimicrobial or antioxidant.

*total.sulfur.dioxie* is the total of free plus bound sulfur dioxide.

The distribution skews to the right, with a mean of 46.47 and a median of 38.

```{r echo=FALSE}
ggplot(rwq, aes(x = total.sulfur.dioxide)) +
  geom_histogram(bins = 30, color = "red") +
  scale_x_log10()
```

A log transformation makes it much more symmetric.


### *sulphates*

```{r sulphates, echo=FALSE}
ggplot(rwq, aes(x = sulphates)) +
  geom_histogram(binwidth = 0.02, color = "red")
```

*sulphates* is the amount of potassium sulphate in the wine, in gram / liter.
It is an additive to increase the sulfur dioxide levels.

```{r echo=FALSE}
ggplot(rwq, aes(x = sulphates)) +
  geom_histogram(bins = 50, color = "red") +
  scale_x_log10()
```

Again the logarithmic transformation makes the distribution more symmetric.


### *density*

```{r density, echo=FALSE}
ggplot(rwq, aes(x = density)) +
  geom_histogram(bins = 50, color = "red")
```

*density* is the density of the wine, in gram / ml.

The distribution looks quite normal, with the mean and median at 0.9967 and 
0.9968 respectively. 

Most of the wines have a density < 1, ie, lighter than water.


```{r echo=FALSE}
rwq$fixed.acidity.log10 <- log10(rwq$fixed.acidity)
rwq$residual.sugar.log10 <- log10(rwq$residual.sugar)
rwq$chlorides.log10 <- log10(rwq$chlorides)
rwq$total.sulfur.dioxide.log10 <- log10(rwq$total.sulfur.dioxide)
rwq$sulphates.log10 <- log10(rwq$sulphates)
rwq$bound.sulfur.dioxide.log10 <- log10(rwq$total.sulfur.dioxide - 
                                        rwq$free.sulfur.dioxide)
rwq$quality <- factor(rwq$quality)
rwq$quality.binned <- revalue(rwq$quality, c("3" = "bad", "4" = "bad",
                                             "5" = "average", "6" = "average",
                                             "7" = "good", "8" = "good"))
```



# Univariate Analysis

The data set contains 1599 observations of 12 variables. One of them is 
*quality*, the focus of our study.

Given how few variables there are, we'll use all the other variables in our
study, in hopes to find relationship with quality.

I have created the logarithmic transformations of *fixed.acidity*, 
*residual.sugar*, *chlorides*, *total.sulfur.dioxide* and *sulphates* for future
use.

A *bound.sulfur.dioxide.log10* is created by taking the logarithm of the 
difference between *total.sulfur.dioxide* and *free.sulfur.dioxide*.

I've also binned *quality* into 3 categories, "bad" for *quality* 3 and 4, 
"average" for 5 and 6, "good" for 7 and 8. This will be used in multivariate
analyses to reduce the number of colors used for better contrast.



# Bivariate Plots Section


### Pair plots

```{r echo = FALSE, fig.height=12, fig.width=20, message=FALSE, warning=FALSE}
rwq %>%
  select(quality, alcohol, fixed.acidity.log10, volatile.acidity, citric.acid,
         pH, residual.sugar.log10, chlorides.log10, free.sulfur.dioxide,
         total.sulfur.dioxide.log10, sulphates.log10, density) %>%
  ggpairs()
```

Let's first take a glance over the paired plots.
From the box plots with *quality*, we can see that *alcohol*, 
*volatile.acidity*,  *citric.acid* seem to show the strongest relationship with
*quality*.

We'll explore these pairs, as well as other pairs with high correlation one by
one. The number below the plots between *quality* and other features are the 
Spearman's correlation coefficient, using *quality* as an ordinal factor.


### *alcohol* and *quality*

```{r alcohol quality, echo=FALSE, message=FALSE}
ggplot(rwq, aes(x = quality, y = alcohol)) +
  geom_jitter(width = 0.4, alpha = 0.2) 

cor(rwq$alcohol, as.numeric(rwq$quality), method = "spearman")
```

We can see quite an obvious positive relationship between *alcohol* and 
*quality*.

The spearman's correlation coefficient, at 0.48, is coherent with the 
observation. 

Of course, this isn't to say that higher alcohol content necessarily makes the 
wine better, otherwise even the worst ports and sherries would trump the best 
wines.

Rather, it could be that better made wines happen to have higher alcohol
content, due to a third factor.


### *volatile.acidity* and *quality*

```{r volatile.acidity quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = volatile.acidity)) +
  geom_jitter(width = 0.4, alpha = 0.2) 

cor(rwq$volatile.acidity, as.numeric(rwq$quality), method = "spearman")
```

Unsurprisingly, *volatile.acidity* shows a negative correlation with *quality*, 
at -0.38.

Acetic acid, with its vinegar-y smell, is quite universally agreed upon as 
unpleasant in wine.


### *citric.acid* and *quality*

```{r citric.acid quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = citric.acid)) +
  geom_jitter(width = 0.4, alpha = 0.2)

cor(rwq$citric.acid, as.numeric(rwq$quality), method = "spearman")
```

The relationship betwen *citric.acid* and *quality* is less clear. The 
coefficient is 0.21.


### *chlorides.log10* and *quality*

```{r chlorides quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = chlorides.log10)) +
  geom_jitter(width = 0.4, alpha = 0.1)
cor(rwq$chlorides.log10, as.numeric(rwq$quality), method = "spearman")
```

Again the relationship here is faint, with spearman's coefficient at 0.19.
I expected it to be stronger as saltiness is certainly a negative trait for 
most people.


### Sulphur dioxides and *quality*

```{r total.sulfur.dioxide quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = total.sulfur.dioxide.log10))  +
  geom_boxplot()

cor(rwq$total.sulfur.dioxide.log10, as.numeric(rwq$quality), method = "spearman")
```

```{r bound.sulfur.dioxide quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = bound.sulfur.dioxide.log10)) +
  geom_boxplot()

cor(rwq$bound.sulfur.dioxide.log10, as.numeric(rwq$quality), method = "spearman")
```

```{r free.sulfur.dioxide quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = free.sulfur.dioxide)) +
  geom_boxplot()

cor(rwq$free.sulfur.dioxide, as.numeric(rwq$quality), method = "spearman")
```

None of the 3 variables we have about sulfur dioxide -- *free.sulfur.dioxide*, 
*bound.sulfur.dioxide* and *total.sulfur.dioxide*, are highly correlated to 
*quality* by themselves. The spearman's coefficient are around 0.2.

But all of them seem to show a parabolic relationship with *quality*, if the 
latter is to be interpreted as a continuous variable.

This may warrant further investigation, or some feature engineering.


### *sulphates* and *quality*

```{r sulphates quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = sulphates)) +
  geom_jitter(width = 0.4, alpha = 0.2)

cor(rwq$sulphates, as.numeric(rwq$quality), method = "spearman")
```

With *sulphates* we see quite a clear positive relationship with quality, with 
spearman's coefficient at 0.38.


### *density* and *quality*

```{r density quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = quality, y = density)) +
  geom_jitter(width = 0.4, alpha = 0.2)

cor(rwq$density, as.numeric(rwq$quality), method = "spearman")
```

The relationship between *density* and *quality* is quite weak.

Also, *density* is highly correlated to many other variables we have, which, in 
turn, are correlated to *quality*. It is doubtful if *density* is still 
meaningful if those other variables are accounted for.


### *citric.acid* and *fixed.acidity*

```{r citric.acid fixed.acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = citric.acid, y = fixed.acidity)) +
  geom_point(alpha = 0.1) + 
  geom_smooth()
```

This is one of the strongly correlated pairs among all the variables we
have. 


### *fixed.acidity.log10* and *pH*

```{r fixed.acidity.log10 pH, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = fixed.acidity.log10, y = pH)) +
  geom_point(alpha = 0.1) + 
  geom_smooth()
```

As one would expect, *fixed.acidity* has a strong negative correlation with 
*pH*.


### *volatile.acidity* and *pH*

```{r volatile.acidity pH, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = volatile.acidity, y = pH)) +
  geom_point(alpha = 0.1) + 
  geom_smooth()
```

A little surprising is that *volatile.acidity* is slightly positively 
correlated with *pH*, at 0.235.


### *alcohol* and *density*

```{r alcohol density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = alcohol, y = density)) +
  geom_point(alpha = 0.1) + 
  geom_smooth() 
```

*Density* is correlated with many of our variables, one of which is *alcohol*.
The negative correlation makes sense, as alcohol is less dense than water.


### *fixed.acidity.log10* and *density*

```{r fixed.acidity density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = fixed.acidity.log10, y = density)) +
  geom_point(alpha = 0.1) + 
  geom_smooth()
```

Correlation coefficient is 0.675. 


### *residual.sugar.log10* and *density*

```{r residual.sugar density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = residual.sugar.log10, y = density)) +
  geom_point(alpha = 0.1) + 
  geom_smooth() 
```

This plot may look a bit quirky, but in the left half of the plot that contains
most of the observations, *density* actually shows a monotonic relationship with 
*residual.sugar* .



# Bivariate Analysis

Among the variables we have, we can see that quality is most strongly correlated 
with alcohol, followed by volatile.acidity and sulphates.
Some other variables with weaker correlation are citric.acid, chlorides and 
bound.sulfur.dioxide.

The acid-related variables are often correlated with each other, and with pH as 
well, which makes sense. The odd one is volatile.acidity, which is negatively
correlated with the other acid volumes and positively with pH.

Density is highly correlated with many variables such as fixed.acidity and
alcohol.

The pair of variables with the strongest correlation is free.sulfur.dioxide
and total.sulfur.dioxide.log10, with a coefficient of 0.714.



# Multivariate Plots Section


### The top 3 features and *quality*

```{r alcohol sulphates quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = alcohol, y = sulphates.log10, color = quality)) +
  geom_jitter(alpha = 1) +
  scale_color_discrete(h = c(240, 0))

```

With two of our variables with the strongest correlation with quality, we can 
see the pattern clearly in this plot. *alcohol* and *sulphates* correlate
positively with *quality*

```{r alcohol volatile.acidity quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq,  aes(x = alcohol, y = volatile.acidity, color = quality)) +
  geom_jitter(alpha = 1) +
  scale_color_discrete(h = c(240, 0))
```

Again we can see that as the *quality* increases from blue to red, we go from the
top left to bottom right in this plot, ie, lower *volatile.acidity* and higher
*alcohol*.


```{r sulphates volatile.acidity quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = sulphates.log10, y = volatile.acidity, color = quality)) +
  geom_jitter(alpha = 1) +
  scale_color_discrete(h = c(240, 0), drop = FALSE) 
```

The patter is less pronounced in this plot as neither of the two variables here
are as strongly correlated to *quality* as *alcohol*. But still the pattern is
visible.


### *alcohol*, *density* and *quality*

```{r density alcohol, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = alcohol, y = density, color = quality.binned)) +
  geom_jitter(alpha = 1) +
  scale_color_discrete(h = c(240, 0)) 


```

In this plot I use the binned quality (4 and below are bad, 5 and 6 average,
7+ are good). We lose some information details but this way the contrast is 
better and the pattern easier to see.

We can see that *density* has a negative relationship with *alcohol*, as alcohol
is less dense than water. And it seems that among wines with the same alcohol
content levels, the relationship between density and quality almost reverses,
ie, the higher the density, the higher the quality.

I'd speculate that the negative correlation between density and quality is 
completely explained by its negative correlation with alcohol.


### Acidities and *pH*

```{r acidities pH, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rwq, aes(x = volatile.acidity, y = fixed.acidity, color = pH)) +
  geom_jitter(alpha = 1, size = 1) +
  scale_color_gradient(low = "red", high = "green") 
```

Lastly we take a look at how *pH* changes with our two acidity variables.
The very pronounced pattern is the negative relationship between *fixed.acidity*
and *pH*. But we also see that *pH* generally increases as *volatile.acidity*
increases.



# Multivariate Analysis

In the previous section we found that *density* has a negative correlation with
*quality*, but here it seems that when *alcohol* is in the equation the 
effect reverses. 

Most other features behave the same way as we saw in the bivariate section.



# Final Plots and Summary


### Plot One

```{r echo=FALSE, fig.height=6, fig.width=10, Plot_One}
ggplot(rwq, aes(x = quality, y = alcohol)) +
  geom_jitter(width = 0.4, alpha = 0.2) +
  ggtitle("Alcohol content / quality") +
  xlab("Quality rating (10 for best)")+
  ylab("Alcohol content by volume (%)")
```

This plot, though simple, is important as *alcohol* has the strongest 
correlation with *quality* among all the features we have.

We can clearly see that the wines with higher alcohol content are rated higher.
The difference is particularly pronounced between categories 6+ and those below
6.


### Plot Two

```{r echo=FALSE, fig.height=6, fig.width=10, Plot_Two}
ggplot(rwq,  aes(x = alcohol, y = volatile.acidity, color = quality)) +
  geom_jitter(alpha = 0.5, size = 2) +
  scale_color_discrete(h = c(240, 0)) +
  labs(title = "Volatile acidity, alcohol content and quality",
       x = "Alcohol content by volume(%)",
       y = "Volatile acidity (acetic acid - gram / liter)",
       color = "Quality rating (10 for best)")

```

This plot shows the two features with the strongest correlation with *quality*,
which means they probably are the best predictors of a wine quality we can 
find in this data set.


### Plot Three

```{r echo=FALSE, fig.height=6, fig.width=10, Plot_Three}
ggplot(rwq, aes(x = alcohol, y = density, color = quality.binned)) +
  geom_jitter(alpha = 0.5, size = 2) +
  scale_color_discrete(h = c(240, 0)) +
  labs(title = "Density / quality relationship explained by alcohol content?",
       x = "Alcohol content by volume(%)",
       y = "Density (gram / ml)",
       color = "Binned quality rating")
```

This plot is interesting in that we can see, contrary to what we've found in 
the bivariate analysis and correlation figures, once *alcohol* is accounted for, 
*density* seems to have a positive relationship with *quality*.



# Reflection

I am somewhat surprised to find that the strongest predictor we have of quality,
at least when used alone, is such a simple, measureable and easily observable 
feature as alcohol content. 

As far as I know, it isn't extremely hard to increase the alcohol content by a
a little when making wine. At least some of the makers who stop the fermentation
early can stop doing that and let it finish?

But then it could be that some third factor causes both a higher alcohol content
and better quality - imagine that better grapes make better wine, and better
grapes are also suited to complete fermentation. 

If there was such a factor, however, we can't find it in our dataset. 
Statistical modeling or machine learning techniques may provide more insights.


